% ─────────────────────────────────────────────────────────────────────────────
\section{Results}
\label{sec:results}
% ─────────────────────────────────────────────────────────────────────────────

\subsection{Anomaly Detection}
\label{subsec:results_ad}

Table~\ref{tab:ad_results} reports 5-fold cross-validation performance for all
classical models on Repr.~B (184 features), the Ensemble variant on
Repr.~F\textsubscript{++} (224 features), and the published DL
benchmarks.\footnote{All deep learning figures in this report are reproduced
from the LMSD preprint \cite{lmsd2025}. That preprint was subsequently
\emph{withdrawn} by its authors, who stated that ``significant methodological
flaws have been identified in the experimental validation and metric-computation
procedures.'' We therefore present these numbers as \emph{indicative} reference
points from the literature, not as validated upper bounds, and we interpret the
classical-vs.-DL gaps accordingly throughout.}

\begin{table}[H]
    \centering
    \caption{Anomaly Detection results (5-fold CV, mean $\pm$ std).
             Best classical result in \textbf{bold}; best DL result in \textit{italics}.}
    \label{tab:ad_results}
    \begin{tabular}{llcccc}
        \toprule
        \textbf{Model} & \textbf{Repr.} & \textbf{F1} & \textbf{AUC} & \textbf{FNR} & \textbf{MCWPM} \\
        \midrule
        \multicolumn{6}{l}{\textit{Classical models (this work)}} \\
        \textbf{Logistic Reg.}  & B (184) & \textbf{0.691$\pm$0.009} & \textbf{0.750$\pm$0.008} & \textbf{0.299$\pm$0.010} & \textbf{0.532$\pm$0.009} \\
        Linear SVM     & B (184) & 0.685$\pm$0.008 & 0.748$\pm$0.007 & 0.321$\pm$0.008 & 0.522$\pm$0.008 \\
        XGBoost        & B (184) & 0.682$\pm$0.010 & 0.764$\pm$0.013 & 0.332$\pm$0.021 & 0.519$\pm$0.022 \\
        Random Forest  & B (184) & 0.637$\pm$0.011 & 0.722$\pm$0.011 & 0.396$\pm$0.024 & 0.473$\pm$0.020 \\
        KNN            & B (184) & 0.531$\pm$0.007 & 0.575$\pm$0.004 & 0.482$\pm$0.014 & 0.373$\pm$0.008 \\
        \textbf{Ensemble}       & F\textsubscript{++} (224) & \textbf{0.709$\pm$0.004} & \textbf{0.782$\pm$0.010} & \textbf{0.287$\pm$0.014} & \textbf{0.552$\pm$0.014} \\
        \midrule
        \multicolumn{6}{l}{\textit{Deep learning benchmarks (paper \cite{convtok2023})}} \\
        \textit{ConvTokMHSA}  & raw & \textit{0.799} & --- & \textit{0.163} & --- \\
        InceptionTime \cite{inceptiontime2020} & raw & 0.784 & --- & 0.196 & --- \\
        MMK Net        & raw & 0.721 & --- & 0.298 & --- \\
        \bottomrule
    \end{tabular}
\end{table}

\subsubsection{Analysis}

\paragraph{Best classical model.}
Logistic Regression with Repr.~B achieves F1\,=\,0.691, AUC\,=\,0.750, and
FNR\,=\,0.299. Its linear decision boundary over the 184-dimensional feature
space is sufficient to capture the global health state of a flight. The
simplicity of LR also makes it the most deployable classical option:
fast inference, no hyperparameter sensitivity to training data, and full
interpretability through its coefficient vector.

\paragraph{Ensemble improvement.}
Adding cross-physical inter-cylinder features (Repr.~F\textsubscript{++}) and
combining LR with an XGBoost component raises F1 to 0.709 ($+$1.8\,pp) and
reduces FNR to 0.287 ($-$1.2\,pp). The MCWPM improvement from 0.532 to 0.552
reflects a meaningful gain in safety-weighted performance. This confirms that
inter-cylinder temperature asymmetry provides information beyond per-sensor
global statistics.

\paragraph{Gap to DL.}
The best DL model (ConvTokMHSA, F1\,=\,0.799) outperforms the best classical
system by \textbf{9.0\,pp in F1} and \textbf{12.4\,pp in FNR}. ConvTokMHSA
applies global self-attention over the full 2{,}048-step sequence, allowing it
to detect subtle whole-flight coherence patterns---such as gradual thermal drift
correlated with electrical anomalies across flight phases---that are invisible
to any fixed-window or global statistic. This gap represents the minimum cost
of discarding temporal structure for AD.

\paragraph{Feature importance.}
Among the 184 Repr.~B features, the five most discriminative sensors by
classification effect size (Cohen's $d$, healthy vs.\ anomalous) are:
\texttt{E1 OilP} ($d = 0.158$),
\texttt{IAS} ($d = 0.097$),
\texttt{E1 FFlow} ($d = 0.087$),
\texttt{AltMSL} ($d = 0.079$), and
\texttt{E1 RPM} ($d = 0.075$).
Oil pressure is the single most discriminative channel, consistent with its
role as the primary engine health indicator and direct sensor for lubrication
system faults.

\subsection{Fault Classification}
\label{subsec:results_fc}

Table~\ref{tab:fc_results} reports FC results on the 5{,}601 anomalous flights
across multiple representations and models, compared to DL benchmarks.

\begin{table}[H]
    \centering
    \caption{Fault Classification results (5-fold CV, F1-macro). Best classical in \textbf{bold}.}
    \label{tab:fc_results}
    \begin{tabular}{llcc}
        \toprule
        \textbf{Model} & \textbf{Repr.} & \textbf{F1 macro} & \textbf{Accuracy} \\
        \midrule
        \multicolumn{4}{l}{\textit{Classical --- Repr.~B baseline}} \\
        XGBoost        & B (184)   & 0.121$\pm$0.005 & 0.379 \\
        Logistic Reg.  & B (184)   & 0.097$\pm$0.009 & 0.102 \\
        Linear SVM     & B (184)   & 0.066$\pm$0.014 & 0.066 \\
        Random Forest  & B (184)   & 0.034$\pm$0.005 & 0.375 \\
        \midrule
        \multicolumn{4}{l}{\textit{Classical --- improved representations}} \\
        \textbf{XGBoost} & \textbf{F (60)} & \textbf{0.227$\pm$0.020} & \textbf{0.404} \\
        \textbf{Oracle K\,=\,4} & \textbf{F (60)} & \textbf{0.420$\pm$0.022} & --- \\
        \midrule
        \multicolumn{4}{l}{\textit{Deep learning benchmarks (paper)}} \\
        \textit{MMK Net} \cite{lmsd2025} & raw & \textit{0.623} & --- \\
        InceptionTime \cite{inceptiontime2020} & raw & 0.563 & --- \\
        ConvTokSWLA    & raw & 0.523 & --- \\
        ConvTokMHSA    & raw & 0.208 & --- \\
        \bottomrule
    \end{tabular}
\end{table}

\subsubsection{Per-Class Analysis}

Table~\ref{tab:perclass} reports the per-class F1 scores for the best classical
model (XGBoost, Repr.~B) to illustrate the long-tail performance distribution.

\begin{table}[H]
    \centering
    \caption{Per-class F1 (XGBoost, Repr.~B). Classes sorted by sample count.}
    \label{tab:perclass}
    \begin{tabular}{lrr}
        \toprule
        \textbf{Fault Category} & \textbf{$n$} & \textbf{F1} \\
        \midrule
        Intake gasket leak / damage           & 2{,}097 & 0.533 \\
        Rocker cover leak / loose / damage    & 1{,}024 & 0.364 \\
        Baffle plug need repair               &   254   & 0.153 \\
        Baffle screw missing / loose          &   211   & 0.139 \\
        Oil cooler need maintenance           &    75   & 0.127 \\
        \midrule
        Engine seal / tube / bolt loose       &    76   & 0.098 \\
        Cylinder crack / fail / need repair   &   108   & 0.055 \\
        Engine run rough                      &   141   & 0.051 \\
        Baffle tie / tie rod loose            &   144   & 0.047 \\
        Engine idle / RPM issue               &    93   & 0.038 \\
        Pilot / in-flight noticed issue       &    75   & 0.000 \\
        \bottomrule
    \end{tabular}
\end{table}

\subsubsection{Analysis}

\paragraph{Repr.~B baseline.}
XGBoost with Repr.~B achieves F1\,=\,0.121, classifying with reasonable accuracy
only the two head classes (intake gasket and rocker cover), which together
account for 55.7\% of anomalous flights. Minority classes are largely
misclassified as the dominant class---a direct consequence of both the long-tailed
distribution and the global statistics discarding the local fault signatures.

\paragraph{Feature selection gain.}
XGB gain-based selection to 60 features (Repr.~F) nearly \textit{doubles} F1
from 0.121 to 0.227 ($+$88\%). This improvement demonstrates that the 124
features removed from Repr.~B were actively harmful---they added noise that
confused the classifier---rather than simply uninformative. The 60 retained
features are dominated by electrical channels (\texttt{amp2}, \texttt{volt2})
and fuel flow extrema (\texttt{E1 FFlow\_min}), highlighting the diagnostic
relevance of transient low values in these channels.

\paragraph{Oracle hierarchical ceiling.}
Grouping the 19 classes into 4 physically coherent groups via Ward clustering
and training a per-group XGBoost raises F1 to 0.420 ($+$96\% over flat XGB
with Repr.~F). This result quantifies the theoretical gain from perfect fault
taxonomy knowledge: if the type of fault were known at the coarse level
(engine operational vs.\ baffle/intake maintenance), the intra-group
classification becomes substantially easier because groups share similar sensor
signatures.

\paragraph{Structural gap to DL.}
Despite the Oracle improvement to 0.420, the best DL model (MMK~Net,
F1\,=\,0.623) still exceeds the classical ceiling by \textbf{20\,pp}. This
residual gap is structural: MMK~Net's micro-kernel convolutions ($k = 1, 3, 5$)
with no pooling extract local temporal signatures at the time step level, while
even the Oracle hierarchical approach is constrained by the information
available in flight-level global statistics. Remarkably, ConvTokMHSA (global
attention, F1\,=\,0.208) performs \textit{worse} than flat XGBoost with Repr.~F
(0.227), confirming that whole-flight context \textit{hurts} FC by injecting
operational noise into the attention weights.

\subsection{End-to-End Diagnosis}
\label{subsec:results_e2e}

\subsubsection{Performance Ladder}

Table~\ref{tab:e2e_ladder} reports the full E2E performance ladder as a function
of increasing oracle knowledge.

\begin{table}[H]
    \centering
    \caption{E2E performance ladder (F1-macro, 20 classes: 0\,=\,healthy, 1--19\,=\,fault type).
             The ``oracle'' prefix denotes perfect Stage~1 or Stage~2 routing.}
    \label{tab:e2e_ladder}
    \begin{tabular}{lcc}
        \toprule
        \textbf{Configuration} & \textbf{F1 macro} & \textbf{$\Delta$ vs.\ Cascade} \\
        \midrule
        Naive baseline (all HEALTHY)            & 0.034 & --- \\
        Cascade LR$\to$XGB (Repr.~B)            & 0.112 & --- \\
        Cascade XGB (Repr.~F)                   & 0.157 & $+$0.045 \\
        Oracle routing $\to$ XGB (Repr.~B)      & 0.165 & $+$0.053 \\
        \midrule
        Var.~A: real AD + Oracle FC (K\,=\,4)  & 0.342 & $+$0.230 \\
        Var.~B: Oracle AD + Oracle FC (K\,=\,4) \textbf{[ML ceiling]} & 0.457 & $+$0.345 \\
        \midrule
        LMSD (ConvTokMHSA + MMK Net) \cite{lmsd2025} & 0.623 & --- \\
        \bottomrule
    \end{tabular}
\end{table}

\subsubsection{Error Decomposition}

For the Cascade LR$\to$XGB pipeline on the 11{,}446 out-of-fold flights,
Table~\ref{tab:error_decomp} decomposes the total errors by type.

\begin{table}[H]
    \centering
    \caption{E2E error decomposition for Cascade LR$\to$XGB (Repr.~B), out-of-fold.}
    \label{tab:error_decomp}
    \begin{tabular}{lrr}
        \toprule
        \textbf{Error type} & \textbf{Count} & \textbf{\% of all flights} \\
        \midrule
        Correct prediction                   & 5{,}538 & 48.4\% \\
        Stage~1 false negative (fault~$\to$~healthy) & 1{,}675 & 14.6\% \\
        Stage~1 false positive (healthy~$\to$~fault) & 1{,}833 & 16.0\% \\
        Stage~2 wrong fault type             & 2{,}400 & 21.0\% \\
        \midrule
        \textbf{Total errors} & \textbf{5{,}908} & \textbf{51.6\%} \\
        \bottomrule
    \end{tabular}
\end{table}

\subsubsection{Gap Analysis}

Three quantities characterize the performance gap relative to the DL upper bound:

\begin{align}
    \Delta_{\text{routing}}  &= F1(\text{Var.~A}) - F1(\text{Cascade}) = 0.342 - 0.112 = +0.230 \label{eq:delta_routing} \\
    \Delta_{\text{AD}}       &= F1(\text{Var.~B}) - F1(\text{Var.~A})  = 0.457 - 0.342 = +0.115 \label{eq:delta_ad} \\
    \Delta_{\text{residual}} &= F1(\text{DL}) - F1(\text{ML ceiling})  = 0.623 - 0.457 = -0.166 \label{eq:delta_residual}
\end{align}

These results reveal three important findings:

\begin{enumerate}
    \item \textbf{FC is the dominant bottleneck.} The routing gain
          $\Delta_{\text{routing}} = +0.230$ (Equation~\ref{eq:delta_routing})
          dwarfs the AD gain $\Delta_{\text{AD}} = +0.115$. Improving fault
          classification has twice the leverage of improving anomaly detection
          for E2E performance.

    \item \textbf{Stage~1 false negatives are the safety-critical failure mode.}
          Of the 5{,}908 total errors, 1{,}675 (28.4\%) are Stage~1 false
          negatives---faulty flights routed to the healthy label. These never
          reach the fault classifier and are the most dangerous error type in
          an operational context.

    \item \textbf{40\% of errors are Stage~2 wrong-type errors.} Stage~2 errors
          (2{,}400, 40.6\% of all errors) indicate correct anomaly detection
          followed by incorrect fault attribution. This is not a safety failure
          (the flight is still flagged), but prevents actionable maintenance
          recommendations. The 0.166 residual gap between the ML ceiling and
          the reported LMSD figure appears difficult to close with global
          statistical features alone; temporal micro-pattern features (as in
          MMK~Net) are the most likely route to narrowing it, though the exact
          magnitude is uncertain given the withdrawal of the benchmark source.
\end{enumerate}

\subsection{Summary Comparison}
\label{subsec:summary_compare}

Table~\ref{tab:final_summary} consolidates the best results across all tasks
and compares them against DL benchmarks.

\begin{table}[H]
    \centering
    \caption{Final performance summary. $\uparrow$\,=\,higher is better;
             $\downarrow$\,=\,lower is better.}
    \label{tab:final_summary}
    \begin{tabular}{llccc}
        \toprule
        \textbf{Task} & \textbf{Best model / method} & \textbf{F1}$\uparrow$ & \textbf{FNR}$\downarrow$ & \textbf{vs.\ DL best} \\
        \midrule
        AD  & LR Repr.~B              & 0.691 & 0.299 & $-$0.108 \\
        AD  & Ensemble Repr.~F\textsubscript{++}  & 0.709 & 0.287 & $-$0.090 \\
        AD  & ConvTokMHSA (DL)        & 0.799 & 0.163 & --- \\
        \midrule
        FC  & XGB Repr.~B (baseline)  & 0.121 & ---   & $-$0.502 \\
        FC  & XGB Repr.~F (selected)  & 0.227 & ---   & $-$0.396 \\
        FC  & Oracle K\,=\,4 (ceiling) & 0.420 & ---  & $-$0.203 \\
        FC  & MMK Net (DL)            & 0.623 & ---   & --- \\
        \midrule
        E2E & Cascade LR$\to$XGB Repr.~B & 0.112 & --- & $-$0.511 \\
        E2E & Cascade XGB Repr.~F    & 0.157 & ---   & $-$0.466 \\
        E2E & Var.~B ML ceiling       & 0.457 & ---   & $-$0.166 \\
        E2E & LMSD (DL)              & 0.623 & ---   & --- \\
        \bottomrule
    \end{tabular}
\end{table}

The results support the hypothesis stated in Section~\ref{sec:intro}: classical
models with global statistical features are \textit{competitive but not
equivalent} to DL for AD ($-$9\,pp gap), and face a substantially larger gap for
FC ($-$20\,pp even at the Oracle ceiling, measured against indicative DL figures).
The most promising path to closing the remaining FC gap appears to be temporal
deep learning models---specifically, locally restricted architectures like
MMK~Net---rather than further feature engineering on global statistics. We stress
that the magnitude of this gap rests on benchmark numbers from a withdrawn
preprint and should be re-validated once corrected results become available.
